---
title: "Analysis of U.S. Severe Weather Events"
author: "Kelsey Herod"
date: "2025-11-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Synopsis**
This analysis uses 2024 data from the U.S. National Oceanic and Atmospheric Administration (NOAA) Storm Events Database, which documents major storms and weather hazards across the United States. The database includes information on when and where events occur, along with estimates of any fatalities, injuries, and property damage. The purpose of this analysis is to support government and municipal managers responsible for preparing for severe weather by identifying the most harmful storm events to the population, determining the frequency of these events across states and months, and identifying which counties are harmed the most.


## **Data Processing**

**Database:** U.S. National Oceanic and Atmospheric Administration (NOAA) Storm Events Database

**Source:**  
https://www.ncei.noaa.gov/pub/data/swdi/stormevents/csvfiles/

**Downloads:**

- **Details:** StormEvents_details-ftp_v1.0_d2024_c20251118.csv.gz  
- **Fatalities:** StormEvents_fatalities-ftp_v1.0_d2024_c20251118.csv.gz  
- **Locations:** StormEvents_locations-ftp_v1.0_d2024_c20251118.csv.gz  


### **How to download and extract the dataset**

1. Copy and paste the source link into your web browser to access the datasets. The link will lead you directly to the NOAA Storm Events Database.

2. Find and click to download the datasets listed above for Details, Fatalities, and Locations.

3. Find the downloaded datasets by going to where your downloads are usually stored.

4. Move the unzipped files to your preferred document location.

5. Right click on each zipped file and select "Extract All..." to utilize its contents.


### **How to load data into R and prep for analysis**

1. Automated package installation and loading
```{r load_packages, echo=TRUE, message=FALSE, warning=FALSE, comment=NA}
required_packages <- c("dplyr", "tidyr", "readr","ggplot2", "knitr")

for(pkg in required_packages) {
  
  if(!require(pkg, character.only = TRUE)) {
    
    cat(paste("Installing", pkg, "...\n"))
    
    install.packages(pkg, repos = "https://cran.r-project.org")
    
    library(pkg, character.only = TRUE)
    
  } else {
    
    cat(paste(pkg, "version", packageVersion(pkg), "loaded\n"))
    
  }
  
}

```

2. Define the folder path
```{r}
folder_path <- "C:/Users/kelse/OneDrive/Documents/DAT511/Final Project"
```

3. Define the file paths for the unzipped CSV files
```{r}
details_file <- file.path(folder_path, "StormEvents_details-ftp_v1.0_d2024_c20251118.csv")
fatalities_file <- file.path(folder_path, "StormEvents_fatalities-ftp_v1.0_d2024_c20251118.csv")
locations_file <- file.path(folder_path, "StormEvents_locations-ftp_v1.0_d2024_c20251118.csv")
```

4. Load the CSV files into R
```{r message=FALSE, warning=FALSE}
details <- read_csv(details_file)
fatalities <- read_csv(fatalities_file)
locations <- read_csv(locations_file)
```

5. Join the datasets by EVENT_ID
```{r message=FALSE, warning=FALSE}
joined_data <- details %>%
  left_join(locations, by = "EVENT_ID") %>%
  left_join(fatalities, by = "EVENT_ID")
```

6. Save the joined data to a new CSV file
```{r}
output_file <- file.path(folder_path, "StormEvents_joined_data.csv")
write_csv(joined_data, output_file)
```

7. Inform the user of output
```{r comment=NA}
message("Joined data saved to: ", output_file)
```

## **Results**

#### Across the United States, which types of events (as indicated in the EVENT_TYPE variable) are most harmful with respect to population health?

Combine total number of injuries and deaths into one column
```{r}
storm <- joined_data
storm$injuries <- storm$INJURIES_DIRECT + storm$INJURIES_INDIRECT
storm$deaths <- storm$DEATHS_DIRECT + storm$DEATHS_INDIRECT
storm$health <- storm$injuries + storm$deaths
```


Create table for storm event rankings
```{r}
event_summary <- storm %>%
  group_by(EVENT_TYPE) %>%
  summarize(
    total_inj = sum(injuries, na.rm = TRUE),
    total_deaths = sum(deaths, na.rm = TRUE),
    total_health_impact = sum(health, na.rm = TRUE)
  )

event_ranking <- event_summary %>%
  arrange(desc(total_health_impact))
```

Isolate top ten most harmful storm events
```{r}
top10_events <- head(event_ranking, 10)
```

Create a bar chart for top ten events
```{r}
ggplot(top10_events, aes(x = reorder(EVENT_TYPE, total_health_impact),
                         y = total_health_impact)) +
  geom_col(color = "black", fill = "black") +
  coord_flip() +
  labs(title = "Top 10 Harmful Storm Events on Population Health",
       x = "Event Type",
       y = "Injuries & Deaths") +
  theme_minimal()
```

#### Across the United States, which types of events happen the most for each state/territory?

Summary count of storm event type per state/territory
```{r}
events_by_state <- joined_data %>%
  group_by(STATE, EVENT_TYPE) %>%
  summarize(count = n())
```

Storm event type with the highest count in each state/territory
```{r}
most_common_event_by_state <- events_by_state %>%
  group_by(STATE) %>%
  filter(count == max(count))
```

Summary table of most frequent storm event types by state/territory
```{r}
kable(
  most_common_event_by_state,
  caption = "Most Frequent Storm Event Type by State/Territory",
  col.names = c("State", "Event Type", "Event Count"),
  align = c("l", "l", "r")
)
```

#### Which types of events are characterized by which months?

Summary table of storm event count per month
```{r}
events_per_month_event <- joined_data %>%
  group_by(MONTH_NAME, EVENT_TYPE) %>%
  summarize(event_count = n(), .groups = "drop") %>%
  mutate(MONTH_NAME = factor(MONTH_NAME,
                             levels = c("January", "February", "March", "April",
                                        "May", "June", "July", "August",
                                        "September", "October", "November", "December"))) %>%
  arrange(MONTH_NAME, desc(event_count))
```

Summary table of most frequent storm event per month
```{r}
top_event_each_month <- events_per_month_event %>%
  group_by(MONTH_NAME) %>%
  arrange(desc(event_count)) %>%
  slice(1)
```

Bar chart of most frequent storm event per month
```{r}
ggplot(top_event_each_month, aes(x = MONTH_NAME, y = event_count, fill = EVENT_TYPE)) +
  geom_col() +
  labs(
    title = "Most Frequent Storm Event Type Per Month",
    x = "Month",
    y = "Event Count",
    fill = "Event Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


#### Which county has been harmed the most by storm events?

Filter county FIPS number for only county rows
```{r}
county_data <- joined_data %>%
  filter(CZ_TYPE == "C")
```

Create unique county FIPS number by including state FIPS number
```{r}
county_data <- county_data %>%
  mutate(
    county_fips = sprintf("%02d%03d", STATE_FIPS, CZ_FIPS)
  )
```

Combine total number of injuries and deaths into one column
```{r}
county_data <- county_data %>%
  mutate(
    injuries_county = INJURIES_DIRECT + INJURIES_INDIRECT,
    deaths_county   = DEATHS_DIRECT + DEATHS_INDIRECT,
    health_county   = injuries_county + deaths_county
  )
```

Total storm harm (injuries and deaths) by county
```{r}
harm_by_county <- county_data %>%
  group_by(county_fips, STATE, CZ_NAME) %>%
  summarise(
    total_injuries = sum(injuries_county, na.rm = TRUE),
    total_deaths   = sum(deaths_county, na.rm = TRUE),
    total_harm     = sum(health_county, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_harm))
```


Bar chart of top ten harmed counties
```{r}
top10_counties <- harm_by_county %>%
  arrange(desc(total_harm)) %>%
  head(10)

ggplot(top10_counties,
       aes(x = reorder(paste(CZ_NAME, STATE, sep = ", "), total_harm),
           y = total_harm)) +
  geom_col(fill = "red") +
  coord_flip() +
  labs(
    title = "Top 10 Most Harmed Counties in the U.S.",
    x = "County",
    y = "Total Injuries & Deaths"
  ) +
  theme_minimal()
```


